The effect of teachers reassigning students to new Cognitive Tutor sections

### 2020-05-26 paper.Rnw
interactions, sec 6.2

%\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
%\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
% \alignauthor
% Adam C Sales\\
%        \affaddr{University of Texas College of Education}\\
%        \affaddr{Austin, TX, USA}\\
%        \email{asales@utexas.edu}
% % 2nd. author
% \alignauthor
% John F Pane\\
%        \affaddr{RAND Corporation}\\
%        \affaddr{Pittsburgh, PA, USA}\\
%        \email{jpane@rand.org}
% }
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

The second condition for propensity score matching is that there are no unmeasured confounders:
\begin{equation}\label{eq:ignore1}
(y^1,y^0)\independent Z|\bm{x},Class
\end{equation}
% for the student-level analysis and
% \begin{equation*}
% (y^1,y^0)\independent \bar{Z}|\bm{\bar{x}}
% \end{equation*}
% for the classroom-level analysis.
Assumption \eqref{eq:ignore1} is well known as the Achilles heel of causal inference outside of RCTs.
\eqref{eq:ignore1} is untestable; its believability depends on what is understood about the process that underlies treatment assignment $Z$, and what covariates are available for control.
In our case, reassignment is poorly understood, and appears highly idiosyncratic \cite{descriptivePaper}.
Fortunately, our study includes a pretest measure, and observational studies controlling for pretest scores tend to perform well, and replicate experimental estimates \cite{cook2008three} \cite{cook2009bias}.
Section \ref{sec:sensitivity} discusses a sensitivity analysis that
relaxes \label{eq:ignore1} and assumes reasonable levels of unmeasured
confounding.

@@

All in all, the propensity score model was:
\begin{equation}\label{eq:propmod1}
\begin{split}
Pr(&Z_i=1|\bm{x}_i,Class_i=j)=\\
&logit^{-1}( \beta_{0state[i]}+beta_{1j[i]} \tilde{w}_i+\beta_2 \bar{w}_{j[i]}+\\
&\beta_3 Black_i+\beta_4 Latinx_i+\beta_5 Male_i+\\
&\beta_6 Freshman_i+\beta_7 SpEd_i+\beta_8 gifted_i+\\
&\beta_9 ESL_i+\beta_{10} FRL_i+\beta_{11} FRLmis_i+\\
&\beta_{12}year_i+\epsilon^{Cls}_{j[i]}+\epsilon^{Teach}_{k[i]}+\epsilon^{Schl}_{l[i]})
\end{split}
\end{equation}
where $logit^{-1}(x)=e^x/(1+e^x)$ is the inverse logit function, $\beta_{0state[i]}$ is a (fixed) intercept for each state in the sample, $\beta_{1j[i]}$ is a random slope for $\tilde{w}_i$, varying by classroom $j$, $FRLmis_i$ is an indicator for missing data in $FRL$ (which was mode-imputed), $year_i=1,2$ is the study year for subject $i$, $\epsilon^{Cls}_{j[i]}$, $\epsilon^{Teach}_{k[i]}$, and $\epsilon^{Schl}_{l[i]}$ are random intercepts for classroom $j$, teacher $k$, and school $l$.
The random intercepts were all modeled as normal with a mean of zero and a standard deviation estimated from the data. The random slope $\beta_{1j[i]}$ was modeled as normal with a mean and standard deviation estimated from the data. $\beta{1j}$ and $\epsilon^{Cls}_j$ were allowed to correlate, while other random effects were modeled as independent.

@@

If reassignment is used differently from classroom to classroom, it
stands to reason that it might have different effects in different
classrooms, as well.
To test that assumption, we fit a multilevel model with random effects
for reassignment, varying by classroom.
The model had the same fixed effects as model underlying the
``Match+Regression'' results described above, as well as random
intercepts for classroom and random slopes for reassignment, varying
by classroom.
That is, it models the effect of reassignment for a student in a
particular classroom as normally distributed with a mean and a
variance estimated from the data.
While estimating the effect of reassignment in any particular
classroom is beyond the scope of our data, this model allows us to estimate the variance
across those effects.

\begin{figure}
\includegraphics[width=\linewidth]{plots/EffectByClassroom.pdf}
\caption{Classroom-specific effects of reassignment from the
  multilevel model described in Section \ref{sec:heterogeneity}.}
\label{fig:trtHet}
\end{figure}

@@

Therefore, while the effect of reassignment was negative in most classrooms, it may
have been positive in some.
This variation could be due to a number of factors, including
differences in the composition of classrooms, but supports the
hypothesis that differences in when and how reassignment is used
lead to differences in its effect.

@@

<<include=FALSE,cache=FALSE>>=
detach(dataInfo)
@
% \section{Acknowledgments}
% This section is optional; it is a location for you
% to acknowledge grants, funding, editing assistance and
% what have you.  In the present case, for example, the
% authors would like to thank Gerald Murray of ACM for
% his help in codifying this \textit{Author's Guide}
% and the \textbf{.cls} and \textbf{.tex} files that it describes.

>>>>

\numberofauthors{2} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Adam C Sales\\
       \affaddr{University of Texas College of Education}\\
       \affaddr{Austin, TX, USA}\\
       \email{asales@utexas.edu}
% 2nd. author
\alignauthor
John F Pane\\
       \affaddr{RAND Corporation}\\
       \affaddr{Pittsburgh, PA, USA}\\
       \email{jpane@rand.org}
}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

The second condition for propensity score matching is that there are no unmeasured confounders:
\begin{equation}\label{eq:ignore1}
(y^1,y^0)\independent Z|\bm{x},Class
\end{equation}

Assumption \eqref{eq:ignore1} is well known as the Achilles heel of causal inference outside of RCTs.
\eqref{eq:ignore1} is untestable; its believability depends on what is understood about the process that underlies treatment assignment $Z$, and what covariates are available for control.
In our case, reassignment is poorly understood, and appears highly idiosyncratic \cite{descriptivePaper}.
Fortunately, our study includes a pretest measure, and observational studies controlling for pretest scores tend to perform well, and replicate experimental estimates \cite{cook2008three} \cite{cook2009bias}.
Section \ref{sec:sensitivity} discusses a sensitivity analysis that
relaxes \ref{eq:ignore1} and assumes reasonable levels of unmeasured
confounding.

@@

All in all, the propensity score model was:
\begin{equation}\label{eq:propmod1}
\begin{split}
log&it\left\{Pr(Z_i=1|\bm{x}_i,Class_i=j)\right\}=\\
& \beta_{0state[i]}+\beta_{1} \tilde{w}_i+\beta_2 \bar{w}_{j[i]}+\\
&\beta_3 Black_i+\beta_4 Latinx_i+\beta_5 Male_i+\\
&\beta_6 Freshman_i+\beta_7 SpEd_i+\beta_8 gifted_i+\\
&\beta_9 ESL_i+\beta_{10} FRL_i+\beta_{11}
FRLmis_i+\beta_{12}year_i+\\
&\gamma_{j[i]}\tilde{w}_i+\epsilon^{Cls}_{j[i]}+\epsilon^{Teach}_{k[i]}+\epsilon^{Schl}_{l[i]}
\end{split}
\end{equation}
where $logit(x)=log(x/(1-x))$ is the logit function,
$\beta_{0state[i]}$ is a (fixed) intercept for each state in the
sample, $FRLmis_i$ is an indicator for missing data in $FRL$ (which
was mode-imputed), and $year_i=1,2$ is the study year for subject $i$.

Finally, $\gamma_{j[i]}$, $\epsilon^{Cls}_{j[i]}$,
$\epsilon^{Teach}_{k[i]}$, and $\epsilon^{Schl}_{l[i]}$ are random
effects.
The subscripts $j$, $k$ and $l$ refer to classroom, teacher, and
school, respectively; the $[i]$ refers to student, so that $j[i]$ is
$i$'s classroom, $k[i]$ is $i$'s teacher, and $l[i]$ is $i$'s school.

$\gamma_{j[i]}$ is a random slope for $\tilde{w}_i$, varying at the
classroom level.
This is essentially an interaction term, allowing the slope for
(classroom centered) pretest scores to vary from one classroom to the
next.
However, unlike standard regression interactions, random slopes are
modeled as being drawn from a normal distribution, with a standard
deviation estimated from the data.
This is a form of regularization, shrinking the classroom-level slopes
towards a common value, and allowing stable estimation even with very
few observations from each classroom \cite{gelmanHill}
\cite{partialPoolingEDM}.
The set of random slopes $\gamma_j$ has a mean of zero---the
average slope across classrooms is the fixed intercept $\beta_1$.
Therefore, the slope for pretest in classroom $j$ is
$\beta_1+\gamma_j$.

$\epsilon^{Cls}_{j}$, $\epsilon^{Teach}_{k}$, and $\epsilon^{Schl}_{l}$
  are random intercepts for classroom, teacher, and school.
These were also modeled as normal with a mean of zero and a standard
deviation estimated from the data.
Including them in the regression accounts the fact
that two students in the same classroom or with the same teacher or in
the same school may be more likely to have the same $Z$---either both
be reassigned or neither---than two students in different classrooms,
with different teachers, or in different schools.

@@

If reassignment is used differently from classroom to classroom, it
stands to reason that it might have different effects in different
classrooms, as well.
To test that assumption, we fit a multilevel model with random effects
for reassignment, varying by classroom.
The model had the same fixed effects as model underlying the
``Match+Regression'' results described above, as well as random
intercepts for classroom and random slopes for reassignment, varying
by classroom.
Formally, the model is:
\begin{equation}\label{eq:heteroReg}
\begin{split}
y_i&=\beta_{0,m[i]}+\beta_1Z_i+\\
&\beta_2SpEd_i+\beta_3frlMIS_i+\beta_4Black_i+\beta_5Hisp_i+\\
&ns^5(pretest_i,\bm{\alpha})+\gamma_{j[i]}Z_i+\epsilon^{Cls}_{j[i]}+\epsilon^{Ind}_i
\end{split}
\end{equation}
where $\beta_{0,m[i]}$ is a fixed intercept for each match, $ns^5(pretest_i,\bm{\alpha})$ is a natural spline for pretest, with five degrees of freedom and coefficient vector $\alpha$, and $\gamma_{j[i]}$, $\epsilon^{Cls}_{j[i]}$, and $\epsilon^{Ind}_i$ are random effects, modeled as normal with mean zero and standard deviation estimated from the data.
Symbols $\alpha$, $\beta$, $\gamma$, and $\epsilon$ do not represent the same quantities as in equation \eqref{eq:propmod1}. 
$\gamma_{j[i]}$ is the random slope for reassignment, varying by classroom; the effect of reassignment in classroom $j$ is estimated as $\hat{\beta}_1+\hat{\gamma}_j$.
That is, $\beta_1$ represents the effect of reassignment, averaged over all classrooms, and $\gamma_j$ represents the difference between classroom $j$'s effect and the average. 
While precisely estimating the effect of reassignment in any particular
classroom is beyond the scope of our data, this model allows us to estimate the variance across those effects, as the variance of $\gamma_j$s.

\begin{figure}
\includegraphics[width=\linewidth]{plots/EffectByClassroom.pdf}
\caption{Classroom-specific effects of reassignment ($\hat{\beta}_1+\hat{\gamma}_j$) from model \eqref{eq:heteroReg}. Error bars represent standard errors.}
\label{fig:trtHet}
\end{figure}

@@

Therefore, while the effect of reassignment was negative, on average, it may
have been positive in some classrooms.

\begin{figure*}
\centering
\includegraphics{plots/heterogeneityRegression.pdf}
\caption{The random effects $\gamma_j$ from model \eqref{eq:heteroReg} (with error bars for one standard error) as a function of classroom-level variance in pretest scores and the proportion of students in a classroom who were ever reassigned. OLS fits are added for interpretation.}
\label{fig:hetRegs}
\end{figure*}

This variation could be due to a number of factors, including
differences in the composition of classroomsand in when and how reassignment is used.
We considered two simple hypotheses about classroom-level predictors of heterogeneous treatment effects.
The first hypothesis was that variance in students' pretest scores within a classroom predicts the effect of reassignment in that classroom.
The idea is that some teachers may use reassignment as a tool to address varying student ability---for instance, they may reassign lagging students to help them keep up with their classmates.
Classrooms with higher variance in pretest scores afford more opportunities for teachers to use this reassignment strategy. 
If the strategy is widely used, and either particularly effective or ineffective at boosting students' posttest scores, there will be a correlation between classroom-level variance in pretest scores and the effect of reassignment.

Our second hypothesis was that the proportion of students in a classroom who have been reassigned may predict classroom-level effects.
The idea here is that in classrooms with low a low proportion of students reassigned, teachers use reassignment in a more targeted fashion, so it may be more beneficial.

Figure \ref{fig:hetRegs} plots random effects $\hat{\gamma}_j$ from model \eqref{eq:heteroReg} as a function of classroom level pretest variance and the proportion of students reassigned, respectively, with simple OLS fits. 
A positive relationship between pretest variance and $\hat{\gamma}$, and a negative relationship between proportion reassigned and $\hat{\gamma}$ are apparent, but with wide standard errors.
To test these hypotheses more formally, we re-fit model \eqref{eq:heteroReg}, adding fixed effects for the variance in pretest scores and proportion reassigned, as main effects and interacted with $Z_i$.
The model reduced the unexplained variance in classroom-level effects from \Sexpr{round(sdEff,2)} to \Sexpr{round(sdEff2,2)}---these variables explained about \Sexpr{round((1-sdEff2^2/sdEff^2)*100)}\% of the unexplained variance in treatment effects.
The coefficient on the interaction beween pretest variance and reassignment---measuring the extent to which pretest variance explains treatment effects---was estimated as \Sexpr{round(classHet['everCPTRUE:pretestVar','Estimate'],2)}, with a 95\% confidence interval of [\Sexpr{paste(round(ciHet['everCPTRUE:pretestVar',],2),collapse=',')}], so the data are compatible with large associations in either direction between pretest variance and treatment effects.
No firm conclusions may be drawn.
The coefficient on the interaction beween proportion reassigned and reassignment---measuring the extent to which classroom proportion reassigned explains treatment effects---was estimated as \Sexpr{round(classHet['everCPTRUE:meanCP','Estimate'],2)}, with a 95\% confidence interval of [\Sexpr{paste(round(ciHet['everCPTRUE:meanCP',],2),collapse=',')}], and a p-value of $p=\Sexpr{round(classHet['everCPTRUE:meanCP','Pr(>|t|)'],2)}$.
This suggests that the effect of reassignment may be lower---more negative---in classrooms in which a higher proportion of students were reassigned.
This aligns with our second hypothesis.

These effect heterogeneity analyses assume \eqref{eq:ignore1}, no unmeasured confounding. 
Unfortunately, we are not aware of methods for sensitivity analysis of the type presented in Section \ref{sec:sensitivity}, applied towards estimates of effect heterogeneity. 
In particular, unobserved confounding may vary by classroom; for instance, the structure of the propensity score match may vary with the proportion of students ever reassigned, since within-classroom matches will be scarce when this proportion is high.
For those reasons, the conclusions in this section should be taken as suggestive and exploratory.

@@

<<include=FALSE,cache=FALSE>>=
detach(dataInfo)
@

\section{Acknowledgments}
This work was partially funded by NSF grant \#DRL-1420374.
